---
title: "Data Exploration"
output:
  pdf_document: default
  word_document: default
date: "2024-05-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
```

```{r loading-in-data, echo = FALSE}
# Loading in all the data files

daily_act <- read.csv("Fitabase_Data_3.12.16-4.11.16/dailyActivity_merged.csv")
heartrate_sec <- read.csv("Fitabase_Data_3.12.16-4.11.16/heartrate_seconds_merged.csv")
hourly_Cal <- read.csv("Fitabase_Data_3.12.16-4.11.16/hourlyCalories_merged.csv")
hourly_Int <- read.csv("Fitabase_Data_3.12.16-4.11.16/hourlyIntensities_merged.csv")
hourly_Step <- read.csv("Fitabase_Data_3.12.16-4.11.16/hourlySteps_merged.csv")
min_Cal <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteCaloriesNarrow_merged.csv")
min_Int <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteIntensitiesNarrow_merged.csv")
min_METs <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteMETsNarrow_merged.csv")
min_Sleep <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteSleep_merged.csv")
min_Step <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteStepsNarrow_merged.csv")
weight_Log <- read.csv("Fitabase_Data_3.12.16-4.11.16/weightLogInfo_merged.csv")
```

```{r row-col-datasets, echo=FALSE}
# Initial data exploration each data set
# counting the numbers of rows and columns in each variable containing the dataset
# daily_act
nrow_daily_act <- nrow(daily_act)
ncol_daily_act <- ncol(daily_act)
uniqueId_daily_act <- n_distinct(daily_act$Id)
# heartrate_sec
nrow_heartrate_sec <- nrow(heartrate_sec)
ncol_heartrate_sec <- ncol(heartrate_sec)
uniqueId_heartrate_sec <- length(unique(heartrate_sec$Id))
# hourly_Cal
nrow_hourly_Cal <- nrow(hourly_Cal)
ncol_hourly_Cal <- ncol(hourly_Cal)
uniqueId_hourly_Cal <- length(unique(hourly_Cal$Id))
# hourly_Int
nrow_hourly_Int <- nrow(hourly_Int)
ncol_hourly_Int <- ncol(hourly_Int)
uniqueId_hourly_Int <- length(unique(hourly_Int$Id))
# hourly_Step
nrow_hourly_Step <- nrow(hourly_Step)
ncol_hourly_Step <- ncol(hourly_Step)
uniqueId_hourly_Step <- length(unique(hourly_Step$Id))
# min_Cal
nrow_min_Cal <- nrow(min_Cal)
ncol_min_Cal <- ncol(min_Cal)
uniqueId_min_Cal <- length(unique(min_Cal$Id))
# min_Int
nrow_min_Int <- nrow(min_Int)
ncol_min_Int <- ncol(min_Int)
uniqueId_min_Int <- length(unique(min_Int$Id))
# min_METs
nrow_min_METs <- nrow(min_METs)
ncol_min_METs <- ncol(min_METs)
id_min_METs <- length(unique(min_METs$Id))
# min_Sleep
nrow_min_Sleep <- nrow(min_Sleep)
ncol_min_Sleep <- ncol(min_Sleep)
uniqueId_min_Sleep <- length(unique(min_Sleep$Id))
# min_Step
nrow_min_Step <- nrow(min_Step)
ncol_min_Step <- ncol(min_Step)
uniqueId_min_Step <- length(unique(min_Step$Id))
# weight_Log
nrow_weight_Log <- nrow(weight_Log)
ncol_weight_Log <- ncol(weight_Log)
uniqueId_weight_Log <- length(unique(weight_Log$Id))
```

# Background Information

The data sets found in this git repo comes from Kaggle ([FitBit Fitness Tracker Data](https://www.kaggle.com/datasets/arashnic/fitbit?resource=download)). In the Kaggle post, the data set comes from a survey via Amazon Mechanical Turk between 03/12/2016 to 05/12/2026. The entry of each data from the users were recorded between early March to early April.

# Rows, Columns, and Users In Each Dataset

The following are number of rows and columns present in the data sets found in "Fitabase_Data_3.12.16-4.11.16" folder.

1.  In the data set, dailyAcvivity_merged.csv there are `r nrow_daily_act` rows, `r ncol_daily_act` columns, and `r uniqueId_daily_act` unique users.

2.  In the data set, heartrate_seconds_merged.csv there are `r nrow_heartrate_sec` rows, `r ncol_heartrate_sec` columns and `r uniqueId_heartrate_sec` unique users.

3.  In the data set, hourlyCalories_merged.csv there are `r nrow_hourly_Cal` rows, `r ncol_hourly_Cal` columns, and `r uniqueId_hourly_Cal` unique users.

4.  In the data set, hourlyIntensities_merged.csv there are `r nrow_hourly_Int` rows, `r ncol_hourly_Int` columns, and `r uniqueId_hourly_Int` unique users.

5.  In the data set, hourlySteps_merged.csv there are `r nrow_hourly_Step` rows, `r ncol_hourly_Step` columns, and `r uniqueId_hourly_Step` unique users.

6.  In the data set, minuteCaloriesNarrow_merged.csv there are `r nrow_min_Cal` rows, `r ncol_min_Cal` columns, and `r uniqueId_min_Cal` unique users.

7.  In the data set, minuteIntensitiesNarrow_merged.csv there are `r nrow_min_Int` rows, `r ncol_min_Int` columns, and `r uniqueId_min_Int` unique users.

8.  In the data set, minuteMETsNarrow_merged.csv there are `r nrow_min_METs` rows, `r ncol_min_METs` columns, and `r id_min_METs` unique users.

9.  In the data set, minuteSleep_merged.csv there are `r nrow_min_Sleep` rows, `r ncol_min_Sleep` columns, and `r uniqueId_min_Sleep` unique users.

10. In the data set, minuteStepsNarrow_merged.csv there are `r nrow_min_Step` rows, `r ncol_min_Step` columns, and `r uniqueId_min_Step` unique users.

11. In the data set, weightLogInfo_merged.csv there are `r nrow_weight_Log` rows, `r ncol_weight_Log` columns, and `r uniqueId_weight_Log` unique users.

# Data Visualization Of Each Dataset

## Daily Activity Dataset

-   id: User id

-   Activity Date: Date of Activity

-   TotalSteps: Total steps taken for the duration of the activity

-   TotalDistance: Total distance traveled for the duration of the activities. The units recorded are in miles could either be in kilometers or miles depending on which option an user chose. The distance recorded in this column comes from a GPS, manual input from the user, or third party apps.

-   TrackerDistance: Total distance traveled recorded by a tracker. The units recorded are in miles could either be in kilometers or miles depending on which option an user chose. The distance recorded in this column comes from the device and not from a GPS, manual input from the user or third party apps.

-   LoggedActivitiesDistance: Logged actitvites distance refers to distance users manually logged into the device.

-   VeryActiveDistance: Very Active distance refers to distance that the device classifies the physical activity as high intensity based on device's proprietary algorithm.

-   ModeratelyActiveDistance: Moderately Active distance refers to distance that the device classifies the physical activity as moderate intensity based on device's proprietary algorithm.

-   LightActiveDistance: Light Active distance refers to distance that the device classifies as "light active" based on device's proprietary algorithm.

-   SedentaryActiveDistance: Sedentary active distance refers to the distance when the user is not considered "active" by the device's propitiatory algorithm.

-   VeryActiveMinutes: Very Active minutes refers to the time (recorded in minutes) that the device classifies the physical activity as high intensity based on device's proprietary algorithm.

-   ModeratelyActiveMinutes: Moderately Active minutes refers to the time (recorded in minutes) that the device classifies the physical activity as moderate intensity based on device's proprietary algorithm.

-   LightActiveMinutes: Light Active minutes refers to the time (recorded in minutes) that the device classifies as "light active" based on device's proprietary algorithm.

-   SedentaryMinutes: Sedentary active minutes refers to the time (recorded in minutes) when the user is not considered "active" by the device's propitiatory algorithm.

-   Calories: Calories is the calories burned during the time which the activity took place.

```{r daily-act-first-five-rows}
# printing the first five rows
head(daily_act, 5)
```

```{r histogram-daily-act}
# Creating the histograms
hist(daily_act$TotalSteps, main = "Histogram of Total Steps", xlab = "Total Steps Taken (Regardless of User Id)")
hist(daily_act$TotalDistance, main = "Histogram of Total Distance", xlab = "Total Distance (Regardless of User Id)")
hist(daily_act$TrackerDistance, main = "Histogram of Tracker Distance", xlab = "Tracker Distance (Regardless of User Id)")
hist(daily_act$LoggedActivitiesDistance, main = "Histogram of Logged Activities Distance", xlab = "Logged Activities Distance (Regardless of User Id)")
hist(daily_act$VeryActiveDistance, main = "Histogram of Very Active Distance", xlab = "Very Active Distance (Regardless of User Id)")
hist(daily_act$ModeratelyActiveDistance, main = "Histogram of Moderately Active Distance", xlab = "Moderately Active Distance (Regardless of User Id)")
hist(daily_act$LightActiveDistance, main = "Histogram of Light Active Distance", xlab = "Light Active Distance (Regardless of User Id)")
hist(daily_act$SedentaryActiveDistance, main = "Histogram of Sedentary Active Distance", xlab = "Sedentary Active Distance (Regardless of User Id)")
hist(daily_act$VeryActiveMinutes, main = "Histogram of Very Active Minutes", xlab = "Very Active Minutes (Regardless of User Id)")
hist(daily_act$FairlyActiveMinutes, main = "Histogram of Fairly Active Minutes", xlab = "Fairly Active Minutes (Regardless of User Id)")
hist(daily_act$LightlyActiveMinutes, main = "Histogram of Lightly Active Minutes", xlab = "Lightly Active Minutes (Regardless of User Id)")
hist(daily_act$SedentaryMinutes, main = "Histogram of Sedentary Minutes", xlab = "Sedentary Minutes (Regardless of User Id)")
hist(daily_act$Calories, main = "Histogram of Calories", xlab = "Calories (Regardless of User Id)")

```

Majority of the histogram has a skewness towards the left-hand side or the right-hand side. In context, the skewness of all histogram aside from calories make sense since each user may have different amount of entries per day and each user's physical activity may vary from each other. For instance, two users on the same day can run different distances. If these distances were extreme then skewness should occur. The histogram for calories (by inspection) forms a distribution that resembles a normal distribution.

## Heart Rate Seconds 

-   id: User id

-   Time: Date and time stamp

-   Value: Heart beat per 5 seconds recorded by the device

```{r heartrate-seconds-first-five-rows}
# printing the first five rows
head(heartrate_sec, 5)
```

```{r histogram-heartrate-seconds}
# Creating the histograms
hist(heartrate_sec$Value, main = "Histogram of Heart Rate (In Seconds)", xlab = "Total Heart Rate (Regardless of User Id)")
```

In the histogram for heart rate, the histogram users heart rate has a left skew. Indicating that a majority of the heart rate is lower and closer to resting heart rate of 60-100 bpm ([What's a normal resting heart rate?](https://www.mayoclinic.org/healthy-lifestyle/fitness/expert-answers/heart-rate/faq-20057979#:~:text=A%20normal%20resting%20heart%20rate%20for%20adults%20ranges%20from%2060,to%2040%20beats%20per%20minute.)).

## Hourly Calories

-   id: User id

-   Time: Date and time stamp

-   Calories: Calories burned per hour

```{r hourly-calories-first-five-rows}
# printing the first five rows
head(hourly_Cal, 5)
```

```{r histogram-hourly-calories}
# Creating the histograms
hist(heartrate_sec$Value, main = "Histogram of Heart Rate (In Seconds)", xlab = "Total Heart Rate (Regardless of User Id)")
```
