---
title: "Data Exploration"
output:
  pdf_document: default
  word_document: default
date: "2024-05-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
```

```{r loading-in-data, echo = FALSE}
# Loading in all the data files

daily_act <- read.csv("Fitabase_Data_3.12.16-4.11.16/dailyActivity_merged.csv")
heartrate_sec <- read.csv("Fitabase_Data_3.12.16-4.11.16/heartrate_seconds_merged.csv")
hourly_Cal <- read.csv("Fitabase_Data_3.12.16-4.11.16/hourlyCalories_merged.csv")
hourly_Int <- read.csv("Fitabase_Data_3.12.16-4.11.16/hourlyIntensities_merged.csv")
hourly_Step <- read.csv("Fitabase_Data_3.12.16-4.11.16/hourlySteps_merged.csv")
min_Cal <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteCaloriesNarrow_merged.csv")
min_Int <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteIntensitiesNarrow_merged.csv")
min_METs <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteMETsNarrow_merged.csv")
min_Sleep <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteSleep_merged.csv")
min_Step <- read.csv("Fitabase_Data_3.12.16-4.11.16/minuteStepsNarrow_merged.csv")
weight_Log <- read.csv("Fitabase_Data_3.12.16-4.11.16/weightLogInfo_merged.csv")
```

```{r row-col-datasets, echo=FALSE}
# Initial data exploration each data set
# counting the numbers of rows and columns in each variable containing the dataset
# daily_act
nrow_daily_act <- nrow(daily_act)
ncol_daily_act <- ncol(daily_act)
uniqueId_daily_act <- length(unique(daily_act$Id))
# heartrate_sec
nrow_heartrate_sec <- nrow(heartrate_sec)
ncol_heartrate_sec <- ncol(heartrate_sec)
uniqueId_heartrate_sec <- length(unique(heartrate_sec$Id))
# hourly_Cal
nrow_hourly_Cal <- nrow(hourly_Cal)
ncol_hourly_Cal <- ncol(hourly_Cal)
uniqueId_hourly_Cal <- length(unique(hourly_Cal$Id))
# hourly_Int
nrow_hourly_Int <- nrow(hourly_Int)
ncol_hourly_Int <- ncol(hourly_Int)
uniqueId_hourly_Int <- length(unique(hourly_Int$Id))
# hourly_Step
nrow_hourly_Step <- nrow(hourly_Step)
ncol_hourly_Step <- ncol(hourly_Step)
uniqueId_hourly_Step <- length(unique(hourly_Step$Id))
# min_Cal
nrow_min_Cal <- nrow(min_Cal)
ncol_min_Cal <- ncol(min_Cal)
uniqueId_min_Cal <- length(unique(min_Cal$Id))
# min_Int
nrow_min_Int <- nrow(min_Int)
ncol_min_Int <- ncol(min_Int)
uniqueId_min_Int <- length(unique(min_Int$Id))
# min_METs
nrow_min_METs <- nrow(min_METs)
ncol_min_METs <- ncol(min_METs)
id_min_METs <- length(unique(min_METs$Id))
# min_Sleep
nrow_min_Sleep <- nrow(min_Sleep)
ncol_min_Sleep <- ncol(min_Sleep)
uniqueId_min_Sleep <- length(unique(min_Sleep$Id))
# min_Step
nrow_min_Step <- nrow(min_Step)
ncol_min_Step <- ncol(min_Step)
uniqueId_min_Step <- length(unique(min_Step$Id))
# weight_Log
nrow_weight_Log <- nrow(weight_Log)
ncol_weight_Log <- ncol(weight_Log)
uniqueId_weight_Log <- length(unique(weight_Log$Id))
```

The following are number of rows and columns present in the data sets found in "Fitabase_Data_3.12.16-4.11.16" folder.

1.  In the data set, dailyAcvivity_merged.csv there are `r nrow_daily_act` rows, `r ncol_daily_act` columns, and `r uniqueId_daily_act` unique users.

2.  In the data set, heartrate_seconds_merged.csv there are `r nrow_heartrate_sec` rows, `r ncol_heartrate_sec` columns and `r uniqueId_heartrate_sec` unique users.

3.  In the data set, hourlyCalories_merged.csv there are `r nrow_hourly_Cal` rows, `r ncol_hourly_Cal` columns, and `r uniqueId_hourly_Cal` unique users.

4.  In the data set, hourlyIntensities_merged.csv there are `r nrow_hourly_Int` rows, `r ncol_hourly_Int` columns, and `r uniqueId_hourly_Int` unique users.

5.  In the data set, hourlySteps_merged.csv there are `r nrow_hourly_Step` rows, `r ncol_hourly_Step` columns, and `r uniqueId_hourly_Step` unique users.

6.  In the data set, minuteCaloriesNarrow_merged.csv there are `r nrow_min_Cal` rows, `r ncol_min_Cal` columns, and `r uniqueId_min_Cal` unique users.

7.  In the data set, minuteIntensitiesNarrow_merged.csv there are `r nrow_min_Int` rows, `r ncol_min_Int` columns, and `r uniqueId_min_Int` unique users.

8.  In the data set, minuteMETsNarrow_merged.csv there are `r nrow_min_METs` rows, `r ncol_min_METs` columns, and `r id_min_METs` unique users.

9.  In the data set, minuteSleep_merged.csv there are `r nrow_min_Sleep` rows, `r ncol_min_Sleep` columns, and `r uniqueId_min_Sleep` unique users.

10. In the data set, minuteStepsNarrow_merged.csv there are `r nrow_min_Step` rows, `r ncol_min_Step` columns, and `r uniqueId_min_Step` unique users.

11. In the data set, weightLogInfo_merged.csv there are `r nrow_weight_Log` rows, `r ncol_weight_Log` columns, and `r uniqueId_weight_Log` unique users.

```{r visual-one}

head(daily_act)

data_dailySteps <- daily_act |>
  group_by(Id, ActivityDate) |>
  filter(Id == 1503960366) |>
  summarise(steps_per_day = sum(TotalSteps))

data_dailySteps |>
  ggplot(aes(x = ActivityDate, y = steps_per_day)) +
  geom_line(aes(group = 1)) +
  geom_point() +
  labs(title = "Steps Between March 25th to April 9th",
      x = "Date (March 25th to April 9th)",
      y = "Steps") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
